---
title: "Evaluating Homelessness as an Independent risk factor for Hepatitis A"
author: "Aman Sahani"
#date: '2022-07-02'
output: 
 prettydoc::html_pretty:
   theme: cayman
   highlight: github
#html_document
#editor_options: 
#  markdown: 
#    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Visualization

Mat no - 12201412

email -
[aman.sahani\@stud.th-deg.de](mailto:aman.sahani@stud.th-deg.de){.email}

# Problem Definition

Evaluating Homelessness as an Independent risk factor for Hepatitis A.
Through this project I want to analyze the spread of hepatitis A along
different states of the United States of America throughout the years
and simultaneously evaluate homelessness as an independent risk factor.

# Overview

Hepatitis A is a vaccine-preventable liver infection caused by the
hepatitis A virus (HAV). HAV is found in the stool and blood of people
who are infected. Hepatitis A is very contagious. It is spread when
someone unknowingly ingests the virus --- even in microscopic amounts
--- through close personal contact with an infected person or through
eating contaminated food or drink.

Symptoms of hepatitis A can last up to 2 months and include fatigue,
nausea, stomach pain, and jaundice.

Transmission / Exposure The hepatitis A virus is found in the stool and
blood of people who are infected:

-   Person-to-person contact

-   Eating contaminated food or drink

Since Hepatitis is closely related to Sanitation I am trying to find the
correlation between Hepatitis A spread and Homelessness. The higher
homelessness proportion usually leads to low sanitation and increased
drug abuse both can then easily spread hepatitis.

# Methods

We first Start the analysis by loading the tycho dataset for Hepatitis A
and performing and Exploratory Data Analysis on the subject.

Installing Packages

```{r}
library(dplyr)
library(tidyverse)
library(highcharter)
library(data.table)
library(ggplot2)
library(hrbrthemes)
library(corrplot)
library(usmap)
library(hrbrthemes)
library(viridis)
library(cowplot)
#library(plyr)
libs <- c("dplyr", "ggplot2", "plotly", "reshape2", "magrittr", "ggthemes", "tidyr", "DT", "lubridate", "stringr", "RColorBrewer", "knitr", "highcharter")
lapply(libs, require, character.only = TRUE)

add.libs <- c("rgeos","rgdal", "raster", "viridis", "gtable", "grid", "haven", "viridisLite")
lapply(add.libs, require, character.only = TRUE)
```

Loading Tycho Acute Hepatitis A dataset and having a brief look at the
heading of the table

```{r}
tychodf = read.csv("usa hepatitis a.csv")
```

Performing EDA on the dataset.

1.  Looking at the head of the dataset

```{r}
head(tychodf)
```

compactly displaying the internal structure of the dataset for closer
analysis

```{r}
str(tychodf)
```

This gives a an accurate information of the number of rows and columns
present including the values and their format. We can decide the columns
to keep and what columns we need to transform.

Based on the above Information - Removing the cols not required and
fixing time to years

```{r}

year <- format(as.Date(tychodf$PeriodEndDate, format="%Y-%m-%d"),"%Y")
tychodf$Year <- year
tychodf <- tychodf %>%
  select(ConditionName,Fatalities,Admin1Name,Year,CountValue)
```

Get a sum of CountValue i.e. the number of recorded Hepatitis A Cases.

```{r}
summary(tychodf$CountValue)
```

This give a little more information about the minimum and max value of
the hepatitis A cases reported. The Median is being 7 and Mean is 26.79
for all years for all states.

Five num sum

```{r}
fivenum(tychodf$CountValue)
```

The five number sum tells us about the following minimum value,
lower-hinge value, median value, upper-hinge value and maximum value

First lets analyse the rise and fall of the hepatitis A cases by
aggregating it according to year for each region. Also removed a few
states which had incomplete data Group by Year and create a heatmap to
understand the rise/fall through time

```{r}
hepaAyearsAgg <- setNames( aggregate(tychodf$CountValue,by=list(tychodf$Admin1Name,tychodf$Year), FUN=sum), c("State","Year","Cases"))
usastatearea <- setNames(data.frame(cbind(state.name,state.region , state.area)),c("State","Region","Area in sqmi"))
usastatearea$State %<>% tolower()
hepaAyearsAgg$State %<>% tolower()
usaalldata <- merge(hepaAyearsAgg,usastatearea, by="State",all.x = TRUE)
usaalldata$`Area in sqmi` <- as.integer(as.character(usaalldata$`Area in sqmi`))
usaalldata <- subset(usaalldata, State!="wyoming" & State!="guam" & State!="new york" & State!="district of columbia" & State!="minnesota" & State!="south dakota" & State !="puerto rico")
head(usaalldata)
```

Heatmap

```{r}
tempdf <- usaalldata
tempdf <- filter(usaalldata,Year >=2007 & Year<=2016)
tempdf %<>% group_by(Region,Year) %>% summarise(`Average No of Cases` = mean(Cases))
ylabels <- c("NorthEast","South","Midwest","West")

p <- ggplot(tempdf,aes(Year, Region, fill=`Average No of Cases`)) + geom_tile()+  scale_fill_gradient(low="white", high="blue") +
  theme_ipsum() + scale_y_discrete(labels = ylabels) + ggtitle("Average number of cases per region over years")
options(warn=-1)
p
```

As we can see the number of cases have only reduced slightly indicating
that the Hepatits A is still a major issue.

Plotting a usa map to visualize the mean number of cases across years
according to each state.

```{r}
temp1 <- usaalldata
#temp %<>% filter(Year == 2015)
temp1 %<>% group_by(State) %>% summarise(mean = mean(Cases)) 
#temp
temp1 <- temp1 %>% arrange(desc(mean))
temp1$mean <- temp1$mean
usastatearea <- setNames(data.frame(cbind(state.name,state.abb)),c("State","state"))
usastatearea$State %<>% tolower()
temp1 <- merge(temp1,usastatearea, by="State",all.x = TRUE)
plot_usmap(data = temp1, values = "mean", color = "red",labels = TRUE) + 
  scale_fill_continuous(low = "white", high = "blue",name = "Mean", label = scales::comma) + 
  theme(legend.position = "right", plot.title = element_text(color = "black", size = 12, face = "bold",hjust = 0.5)) + ggtitle("USA States with Mean of Hepatits A cases from 2007 to 2016")
```

We get a gist of states who have been relatively on the higher side for
the number of Hepatitis A cases aggregated over the years.

In order to calculate Correlation let's load all the homelessness and
population dataset.

Lets load homelessness and population dataset of 2007 to 2016

```{r}
dfhomeless = read.csv("2007-2016-Homelessnewss-USA.csv")
year <- format(as.Date(dfhomeless$Year, format="%d/%m/%Y"),"%Y")
dfhomeless$Year <- year
dfhomeless$Count <- as.integer(as.character(dfhomeless$Count))
dfhomeless[is.na(dfhomeless)] <- 0
dfhomeless <- dfhomeless %>%
  select(State,Measures,Year,Count)
#head(dfhomeless)

#Renaming state code to actual State Name
old_names <- c("AK","CA","DC","GA","ID","KY","ME","MS","ND","NJ","OH","PR","SD","VA","WI","AL","CO","DE","HI","IL","LA","MI","MT","NE","NM","OK","RI","TN","VT","WV","AR","CT","FL","IA","IN","MA","MN","NC","NH","NV","OR","SC","TX","WA","WY","AZ","KS","MD","MO","NY","PA","UT","GU")

new_names <- c("Alaska","California","District of Columbia","Georgia","Idaho","Kentucky","Maine","Mississippi","North Dakota","New Jersey","Ohio","Puerto Rico","South Dakota","Virginia","Wisconsin","Alabama","Colorado","Delaware","Hawaii","Illinois","Louisiana","Michigan","Montana","Nebraska","New Mexico","Oklahoma","Rhode Island","Tennessee","Vermont","West Virginia","Arkansas","Connecticut","Florida","Iowa","Indiana","Massachusetts","Minnesota","North Carolina","New Hampshire","Nevada","Oregon","South Carolina","Texas","Washington","Wyoming","Arizona","Kansas","Maryland","Missouri","New York","Pennsylvania","Utah","Guam")

naming_key <- setNames(object = old_names, nm = new_names)

dfhomeless <- dfhomeless %>%
    mutate(State = factor(State, levels = old_names, 
              labels = new_names))
#head(dfhomeless)
dfhomeless <- dfhomeless[complete.cases(dfhomeless),]
dfhomeless <- setNames( aggregate(dfhomeless$Count,by=list(dfhomeless$State,dfhomeless$Year), FUN=sum), c("State","Year","HomelessCount"))

#dfhomeless %<>% group_by(State,Year) %>% summarise(mean = mean(Count))
#dfhomeless[4597,]
#new_DF <- dfhomeless[rowSums(is.na(dfhomeless)) > 0,]
#new_DF
dfhomeless$State %<>% toupper()
head(dfhomeless)

```

Loading Population Dataset

```{r}
popall2010_19 <- read.csv("nst-est2019-alldata.csv")
popall2010_19 <- setNames(popall2010_19[, c("NAME","POPESTIMATE2010","POPESTIMATE2011","POPESTIMATE2012","POPESTIMATE2013","POPESTIMATE2014","POPESTIMATE2015","POPESTIMATE2016")],c("NAME","2010","2011","2012","2013","2014","2015","2016"))
popalldf2000_10 <- read.csv("st-est00int-01.csv")
popalldf2000_10 <- setNames(popalldf2000_10[,c("Geographic.Area","X2001","X2002","X2003","X2004","X2005","X2006","X2007","X2008","X2009")],c("NAME","2001","2002","2003","2004","2005","2006","2007","2008","2009"))

popAllYrsdf <- merge(popalldf2000_10,popall2010_19, by="NAME")
nm1 <- c("2001","2002","2003","2004","2005","2006","2007","2008","2009") 
popAllYrsdf[nm1] <- lapply(popAllYrsdf[nm1], gsub, pattern = ",", replacement = "")
cols.num <- c("2001","2002","2003","2004","2005","2006","2007","2008","2009")
popAllYrsdf[cols.num] <- sapply(popAllYrsdf[cols.num], as.integer)
#sapply(popAllYrsdf, class)
#popAllYrsdf

pop2007_2016df <- popAllYrsdf[,c("NAME","2007","2008","2009","2010","2011","2012","2013","2014","2015","2016")]

pop2007_2016df <- setNames(cbind(pop2007_2016df[1], stack(pop2007_2016df[2:11])),c("State", "PopulationCount","Year"))
pop2007_2016df <- pop2007_2016df[,c(1,3,2)]
pop2007_2016df$State %<>% toupper()

pop2007_2016df$State %<>% tolower()
#tychodf
popWithHepaAcases <- setNames( aggregate(tychodf$CountValue,by=list(tychodf$Admin1Name,tychodf$Year), FUN=sum), c("State","Year","Hepatitis A cases"))
popWithHepaAcases <- subset(popWithHepaAcases, Year >=2007 & Year <= 2016)
popWithHepaAcases$State %<>% tolower()

tempdf <- right_join(pop2007_2016df,popWithHepaAcases, by=c("State"="State","Year"="Year"))
dfhomeless$State %<>% tolower()
alldatadf <- right_join(tempdf,dfhomeless, by=c("State"="State","Year"="Year"))

alldatadf$CasesPer10000 <- (alldatadf$`Hepatitis A cases`/alldatadf$PopulationCount)*10000
alldatadf$HomelessnessPer10000 <- (alldatadf$HomelessCount /alldatadf$PopulationCount)*10000
#alldatadf
usastatearea <- setNames(data.frame(cbind(state.name,state.region , state.area)),c("State","Region","Area in sqmi"))
usastatearea$State %<>% tolower()
combineddf <- merge(alldatadf,usastatearea, by="State",all.x = TRUE)
combineddf$`Area in sqmi` <- as.integer(as.character(combineddf$`Area in sqmi`))
combineddf <- subset(combineddf, State!="wyoming" & State!="guam" & State!="new york" & State!="district of columbia" & State!="minnesota" & State!="south dakota" & State !="puerto rico" & State!="north dakota")
combineddf$Density <- combineddf$PopulationCount/combineddf$`Area in sqmi`
head(combineddf)


```

To identify the correlation between various factors we will use the
correlation matrix. Corellation matrix

```{r}
temp1 <- combineddf
#temp %<>% filter(Year == 2015)
temp1 %<>% select(`Hepatitis A cases`,PopulationCount,HomelessCount,CasesPer10000,HomelessnessPer10000,Density)
res <- cor(temp1, use="pairwise.complete.obs")
round(res, 2)
```

From the table we can identify various correlations but we will plot a
correation graph in order to easily visualize correlations

```{r}
corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45) 
```

We can visualize that Homelessness, Population and Density all have
slight correlation with Hepatitis A cases and the highest correlation
exists between Hepatitis A cases and Population Count.

In order to view the relation between Homelessness and Hepatitis A cases
I will plot a bar graph including mean number of cases for each state
aggregated throughout the years.

```{r}
temp <- combineddf
#temp %<>% filter(Year == 2015)
temp %<>% group_by(State) %>% summarise(mean = mean(`Hepatitis A cases`), meanHomeless = mean(HomelessCount)) 
#temp
temp <- temp %>% arrange(desc(mean))



bar1 <- (highchart() %>%
           hc_title(text ="Hepatitis A and Homelessness Mean Comparison",align="center") %>%
           hc_xAxis(categories = temp$State, title = list(text="State")) %>%
           hc_yAxis(title = list(text="Number of Cases")) %>%
           hc_add_series(name = "Mean Number of Hepatitis Cases", data= temp$mean, stack = "Cases") %>%
           hc_add_series(name = "Mean Number of Homeless", data = temp$meanHomeless/10, stack = "Homeless") %>%
           hc_chart(type = "column",
           options3d = list(enabled = TRUE, beta = 15, alpha = 15)) %>%
           #hc_add_series(temp,"line",hcaes(x=temp$State,y=temp$meanHomeless)) %>%
           hc_add_theme(hc_theme_538())
)
bar1

```

From the above graph it looks like the states having higher homelessness
population have higher Hepatitis A cases such as California, Florida,
Arizona and states having lower homelessness population have lower
hepatitis A cases such as Vermont, Delware etc.

However here we have not take the population count into account. Lets
make the same graph with Homelessness per 100000 and Cases per 100000
for accurate representation.

```{r}
temp <- combineddf
temp %<>% filter(Year == 2016)
temp %<>% group_by(State) %>% summarise(mean = mean(CasesPer10000), meanHomeless = mean(HomelessnessPer10000)) 
#temp
temp <- temp %>% arrange(desc(mean))


bar1 <- (highchart() %>%
           hc_title(text ="Hepatitis A and Homelessness per Population Comparison ",align="center") %>%
           hc_xAxis(categories = temp$State, title = list(text="State")) %>%
           hc_yAxis(title = list(text="Number of Cases per 10000")) %>%
           hc_add_series(name = "Mean Number of Hepatitis Cases per 10000", data= temp$mean, stack = "Cases") %>%
           hc_add_series(name = "Mean Number of Homeless per 1000", data = temp$meanHomeless/10, stack = "Homeless") %>%
           hc_chart(type = "column",
           options3d = list(enabled = TRUE, beta = 15, alpha = 15)) %>%
           #hc_add_series(temp,"line",hcaes(x=temp$State,y=temp$meanHomeless)) %>%
           hc_add_theme(hc_theme_538())
)
bar1

```

The same results are now inconclusive Alaska which were at the last in
previous graph having low homelessness and low hepatitis A cases are now
above having higher number of homelessness but having low number of
hepatitis A cases.

Hawaii can be considered as an Outlier and still there are many states
with higher number of homelessness but having low number of hepatitis A
cases.

We will no take 4 states and try to calculate the relation between
hepatitis A cases, Homelessness and Density over the years keeping
population as a factor.

Two states with Highest mean number of Hepatitis A cases

```{r}

temp4charts <- combineddf
#temp4charts %<>% filter(State=='california')
temp4charts <- temp4charts %>% select(State,Year,CasesPer10000,HomelessnessPer10000,Density)
temp4charts <- setNames(temp4charts,c("State","Year","CasesPer100000","HomelessnessPer10000","Density"))
temp4charts$Density <- temp4charts$Density/10
temp4charts$CasesPer100000 <- temp4charts$CasesPer100000*10
tempcachart <- temp4charts %<>% filter(State=='california')
#tempflchart
df <- melt(tempcachart%>% select(Year,CasesPer100000,HomelessnessPer10000,Density), id.vars='Year')
#df
ca <- df %>%
  ggplot( aes(x=Year, y=value, group=variable, color=variable)) +geom_line() + 
    theme_ipsum() +
    ylab("Count")

temp4charts <- combineddf
#temp4charts %<>% filter(State=='california')
temp4charts <- temp4charts %>% select(State,Year,CasesPer10000,HomelessnessPer10000,Density)
temp4charts <- setNames(temp4charts,c("State","Year","CasesPer100000","HomelessnessPer10000","Density"))
temp4charts$Density <- temp4charts$Density/10
temp4charts$CasesPer100000 <- temp4charts$CasesPer100000*10
tempflchart <- temp4charts %<>% filter(State=='florida')
#tempflchart
df <- melt(tempflchart%>% select(Year,CasesPer100000,HomelessnessPer10000,Density), id.vars='Year')
#df
fl <- df %>%
  ggplot( aes(x=Year, y=value, group=variable, color=variable)) +geom_line()  + theme_ipsum_rc() +
    ylab("Count")
options(warn=-1)

plot_grid(ca, fl, labels=c("California", "Florida"), ncol = 1, nrow = 2)
```

Two states having higher number of homelessness but lower number of
Hepatitis A cases.

```{r}

temp4charts <- combineddf
#temp4charts %<>% filter(State=='california')
temp4charts <- temp4charts %>% select(State,Year,CasesPer10000,HomelessnessPer10000,Density)
temp4charts <- setNames(temp4charts,c("State","Year","CasesPer100000","HomelessnessPer10000","Density"))
temp4charts$Density <- temp4charts$Density/10
temp4charts$CasesPer100000 <- temp4charts$CasesPer100000*10
tempcachart <- temp4charts %<>% filter(State=='vermont')
#tempflchart
df <- melt(tempcachart%>% select(Year,CasesPer100000,HomelessnessPer10000,Density), id.vars='Year')
#df
ca <- df %>%
  ggplot( aes(x=Year, y=value, group=variable, color=variable)) +geom_line() +
    theme_ipsum() +
    ylab("Count")

temp4charts <- combineddf
#temp4charts %<>% filter(State=='california')
temp4charts <- temp4charts %>% select(State,Year,CasesPer10000,HomelessnessPer10000,Density)
temp4charts <- setNames(temp4charts,c("State","Year","CasesPer100000","HomelessnessPer10000","Density"))
temp4charts$Density <- temp4charts$Density/10
temp4charts$CasesPer100000 <- temp4charts$CasesPer100000*10
tempflchart <- temp4charts %<>% filter(State=='oregon')
#tempflchart
df <- melt(tempflchart%>% select(Year,CasesPer100000,HomelessnessPer10000,Density), id.vars='Year')
#df
fl <- df %>%
  ggplot( aes(x=Year, y=value, group=variable, color=variable)) +geom_line() +
    theme_ipsum() +
    ylab("Count")
options(warn=-1)
plot_grid(ca, fl, labels=c('Vermont', 'Oregon'), ncol = 1, nrow = 2)
```

From the above 4 graphs the rise and fall of homelessness is not
directly proportional to the rise and fall of Hepatitis A cases.

Even thought there is slight correlation between Homelessness and
Hepatitis A it is mostly attributed to the population Count. After
taking the population as a variable we found out that there is no direct
correlation between these two.

# Result

Although there seems to be a huge correlation between the number of
hepatitis A cases and homelessness however most of that correlation does
not apply once we take population into account.

\

There is slight correlation between hepatitis A cases per 1000 and
homelessness per 10000 however homelessness is not a direct factor for
hepatitis A and the spread is generally attributed to sanitation and the
recent hepatitis outbreak in the USA is potentially linked with fresh
organic strawberries branded as FreshKampo and HEB

\
